False Discovery Rates, A New Deal

### 2014-10-29 talks/UNC2014.Rmd
added UNC talk draft

% False Discovery Rates, A New Deal
% Matthew Stephens
% 2014/7/8

```{r, include=FALSE}
require("qvalue")
require("ashr")
require("ggplot2")
```

@@

# Before we get started: Let's get Organized!

@@

# So what can you do?

@@

# What are these repository things?

@@

# What is knitr?

@@

# What is Reproducible Research?

@@

# Why is reproducibility important?

@@

# More on git, github, knitr, reproducibility

@@

# The Canonical Genomics Experiment

@@

# FDR, local fdr, and q values

@@

```{r, echo=FALSE}
  #simple simulated example
  ncz = 100 # number of bins in z score histograms
  set.seed(111)
  hh.betahat = rnorm(1000,0,2)
  hh.sebetahat = 1
  hh.zscore = hh.betahat/hh.sebetahat
  hh.pval = pchisq(hh.zscore^2,df=1,lower.tail=F)
  hh.ash = ash(hh.betahat,hh.sebetahat, method="fdr")
  hh.q = qvalue(hh.pval)
```

@@

# Example: FDR estimation

@@

# Is this an important problem?

@@

# Problem 1: The Zero Assumption (ZA)

@@

# Implied distribution of $p$ values under $H_1$

@@

# Implied distribution of Z scores under alternative

@@

# FDR problem 2: different measurement precision

@@

# Example: Mouse Heart Data

@@

load("../scratchwork/Smemo_qbinom_zdat.RData")
# two-sided test
  ttest.pval = function(t, df) {
    pval = pt(t, df = df, lower.tail = T)
    pval = 2 * ifelse(pval < 0.5, pval, 1 - pval)
    return(pval)
}

@@

highxp = xx>1000 # select high expressed genes
pval.high = pval[highxp]
qval.high = qvalue(pval.high)

zdat.ash = ash(zdat[3,],zdat[4,],df=2,method="fdr",mixcompdist="halfuniform")
zdat.ash.high = ash(zdat[3,highxp],zdat[4,highxp],df=2,method="fdr",mixcompdist="halfuniform")

@@

# Mouse Data: Counts vary considerably across genes

@@

# Lower count genes, less power

@@

# Higher count genes, more power
```{r, echo=FALSE, fig.height=4,fig.cap=""}
plot_FDReg_hist(pval.high,pi0=qval.high$pi0,ylim=c(0,4))
```

# Low-count genes dilute signal at high-count genes
```{r, echo=FALSE, fig.height=4,fig.cap=""}
plot_FDReg_hist(pval,pi0=qval$pi0,ylim=c(0,4))
```

# FDR problem 2: low count genes add noise, increase q values
```{r,echo=FALSE, fig.height=4,fig.cap=""}
plot(qval.high$qval,qval$qval[highxp],xlab="q values from high-count gene analysis", ylab="q values from all gene analysis", main="q values for high count genes",xlim=c(0,.5),ylim=c(0,.5))
abline(a=0,b=1,col=2)
```

# Problems: Summary

@@

# FDR via Empirical Bayes

@@

# FDR: A New Deal

@@

# FDR - A New Deal

@@

# Illustration: $g$ a mixture of 0-centered normals

@@

# Illustration: $g$ a mixture of 0-anchored uniforms

@@

# Adaptive Shrinkage

@@

# Recall Problem 1: distribution of alternative Z values multimodal
```{r, echo=FALSE,fig.height=4,fig.cap=""}
nullalthist(hh.zscore,hh.fdrtool$lfdr,main="qvalue")  
```

# Problem Fixed: distribution of alternative Z values unimodal
```{r, echo=FALSE,fig.height=4,fig.cap=""}
nullalthist(hh.zscore,hh.ash$lfdr,main="ash")
```

# Example: FDR estimation

@@

# Recall Problem 2: low count genes add noise, increase q values
```{r,echo=FALSE, fig.height=4,fig.cap=""}
plot(qval.high$qval,qval$qval[highxp],xlab="q values from high-count gene analysis", ylab="q values from all gene analysis", main="q values for high count genes",xlim=c(0,.5),ylim=c(0,.5))
abline(a=0,b=1,col=2)
```

# Problem Fixed: incorporating precision reduces influence of low-count genes
```{r,echo=FALSE, fig.height=4,fig.cap=""}
plot(zdat.ash.high$qvalue,zdat.ash$qvalue[highxp],xlab="q values from high-count gene analysis", ylab="q values from all gene analysis", main="q values for high count genes",xlim=c(0,.5),ylim=c(0,.5))
abline(a=0,b=1,col=2)
```

# A new problem: an embarrassment of riches

@@

# Perhaps we didn't really understand the question?

@@

# The False Sign Rate

@@

# Even with many signals, large $p$ values have high lfsr

@@

# Advantages of False Sign Rate

@@

# Simulated example: $\pi_0$ not identifiable.

@@

# Simulated Example: so fdr not identifiable

@@

# Simulated Example: fsr much more identifiable

@@

# Summary

@@

# Other Applications

@@

# Next steps?

@@

# Thanks

@@

- Including Ester Pantaleo, Scott Powers, Mengyin Lu, Sen Tian, Wei Wang, Zhengrong Xing.

@@

# Pandoc Command used

@@

# Some odd things in the data

```{r, echo=FALSE,fig.height=3}
  plot(zdat[3,],zdat[4,],ylab="standard error",xlab="beta-hat")
  dd[tail(order(zdat[3,])),]
```

# A technicality

- Suppose you estimate $\Pr(\beta_j<0)=0.98$,  $\Pr(\beta_j>0)=0.01$, $\Pr(\beta_j=0) = 0.01$. 
- Should you declare an fdr of 0.01 or 0.02?
- Maybe fsr makes more sense anyway?

# Adaptive Shrinkage of point estimates

- Recall idea: amount of shrinkage depends on measurement precision, $s_j$.

# Adaptive Shrinkage of point estimates

```{r, fig.height=4,echo=FALSE,fig.cap=""}
  res=data.frame(betahat =zdat[3,], Estimate=zdat.ash$PosteriorMean,highxp=highxp)
  cbbPalette <- c("#000000", "#D55E00", "#CC79A7")
  labels = c('betahat','Estimate')
  breaks = c("betahat","Estimate")

  pp= ggplot(data=res,aes(betahat,Estimate,col=as.factor(highxp))) +geom_point(shape=1) +
      geom_abline(colour = "black") +
        xlab("betahat") +
        ylab("shrunk betahat")

  print(pp +scale_y_continuous(limits=c(-2,2)) +
          scale_x_continuous(limits=c(-2,2))  +
           scale_colour_manual(values=cbbPalette,breaks=breaks,labels=labels) +
          coord_equal(ratio=1))

#plot(zdat[3,],zdat.ash$PosteriorMean,main="Less shrinkage for high-count genes (red)",xlab="betahat",ylab="shrunk betahat",xlim=c(-2,2),ylim=c(-2,2), col=1+highxp)
#abline(a=0,b=1,col=2)
```

# Shrinkage is adaptive to information

Need to fix counts.associate to use fdr method in ash

```{r, echo=FALSE,fig.height=4}
#hist(zdat[3,])
#plot(zdat[3,],zdat.ash$PosteriorMean,xlim=c(-0.2,0.2))
#points(zdat[3,16677],zdat.ash$PosteriorMean[16677],col=2,pch=16)
#points(zdat[3,16079],zdat.ash$PosteriorMean[16079],col=2,pch=16)
#x[16677,]
#x[16079,]
```

```{r, echo=FALSE}
#temp = counts.associate(cc,c(-1,1,-1,1),1)
#temp.tscore = temp$zdat[3,]/temp$zdat[4,]
#temp.pval = ttest.pval(temp.tscore,df=2)
#plot(temp$zdat.ash$localfdr,temp.pval)
#identify(temp$zdat.ash$localfdr,temp.pval)
```

```{r, echo=FALSE,fig.height=3,fig.cap=""}
plot(zdat.ash$lfdr,pval,ylab="p value", xlab="ASH local fdr")
```

# Shrinkage is adaptive to information

```{r, echo=FALSE,fig.height=3,fig.cap=""}
plot(zdat.ash$lfdr,pval,ylab="p value", xlab="ASH local fdr")
points(zdat.ash$lfdr[15325],pval[15325],col=2,pch=16)
points(zdat.ash$lfdr[16123],pval[16123],col=2,pch=16)
```

# Shrinkage is adaptive to information

```{r, echo=FALSE}
dd = read.table(paste0("../../stat45800/data/nobrega/expression/counts.txt"), header = TRUE)
xx = rowSums(dd[,2:5])
dd = dd[xx>0,]
xx = xx[xx>0]
cbind(dd[,1:5],pval,zdat.ash$lfdr)[c(15325,16123),]
```

# Recall FDR problem 1: q values increased by low count genes
```{r,echo=FALSE, fig.height=4,fig.cap=""}
plot(qval.high$qval,qval$qval[xx>1000],xlab="q values from high-count gene analysis", ylab="q values from all gene analysis", main="q values for high count genes",xlim=c(0,.5),ylim=c(0,.5))
abline(a=0,b=1,col=2)
```

# ASH q values more robust to inclusion of low count genes

```{r, echo=FALSE,fig.height=3,fig.cap=""}
plot(zdat.ash.high$qvalue,zdat.ash$qvalue[xx>1000],xlab="q values from high only",ylab="q values from all data")
abline(a=0,b=1,col=2,lwd=2)
```

Compare fitted $f(\beta)$, both estimating $\pi_0$ and fixing $\pi_0=0$.
```{r, echo=FALSE,fig.height=4,fig.cap=""}
  x=seq(-4,4,length=100)
  plot(cdf.ash(hh.ash.fdr,x),col=2,type="l",lwd=3,ylab="cdf")
  lines(x,cdf.ash(hh.ash.shrink,x)$y,col=3,lwd=3,lty=2)
```

>>>>

---
title: "False Discovery Rates, A New Deal"
author: Matthew Stephens
date: October 30, 2014
output: ioslides_presentation
---

```{r, include=FALSE}
require("qvalue")
require("ashr")
require("ggplot2")
require("knitr")
```

@@

## Before we get started: Let's get Organized!

@@

## So what can you do?

@@

## What are these repository things?

@@

## What is knitr?

@@

## What is Reproducible Research?

@@

## Why is reproducibility important?

@@

## More on git, github, knitr, reproducibility

@@

## The Canonical Genomics Experiment

@@

## FDR, local fdr, and q values

@@

```{r, echo=FALSE}
  library("ashr") #simple simulated example
  ncz = 100 ## number of bins in z score histograms
  set.seed(111)
  hh.betahat = rnorm(1000,0,2)
  hh.sebetahat = 1
  hh.zscore = hh.betahat/hh.sebetahat
  hh.pval = pchisq(hh.zscore^2,df=1,lower.tail=F)
  hh.ash = ash(hh.betahat,hh.sebetahat, method="fdr")
  hh.q = qvalue(hh.pval)
```

@@

## Example: FDR estimation

@@

## Is this an important problem?

@@

## Problem 1: The Zero Assumption (ZA)

@@

## Implied distribution of $p$ values under $H_1$

@@

## Implied distribution of Z scores under alternative

@@

## FDR problem 2: different measurement precision

@@

## Example: Mouse Heart Data

@@

load("../scratchwork/Smemo_qbinom_zdat.RData")
## two-sided test
  ttest.pval = function(t, df) {
    pval = pt(t, df = df, lower.tail = T)
    pval = 2 * ifelse(pval < 0.5, pval, 1 - pval)
    return(pval)
}

@@

highxp = xx>1000 ## select high expressed genes
pval.high = pval[highxp]
qval.high = qvalue(pval.high)

zdat.ash = ash(zdat[3,],zdat[4,],df=2,method="fdr",mixcompdist="halfuniform")
zdat.ash.high = ash(zdat[3,highxp],zdat[4,highxp],df=2,method="fdr",mixcompdist="halfuniform")
zdat.ash.ES = ash(zdat[3,],zdat[4,],df=2,method="fdr",mixcompdist="halfuniform",model="ES")

@@

## Mouse Data: Counts vary considerably across genes

@@

## Lower count genes, less power

@@

## Higher count genes, more power
```{r, echo=FALSE, fig.height=4,fig.cap=""}
plot_FDReg_hist(pval.high,pi0=qval.high$pi0,ylim=c(0,4))
```

## Low-count genes dilute signal at high-count genes
```{r, echo=FALSE, fig.height=4,fig.cap=""}
plot_FDReg_hist(pval,pi0=qval$pi0,ylim=c(0,4))
```

## FDR problem 2: low count genes add noise, increase q values
```{r,echo=FALSE, fig.height=4,fig.cap=""}
plot(qval.high$qval,qval$qval[highxp],xlab="q values from high-count gene analysis", ylab="q values from all gene analysis", main="q values for high count genes",xlim=c(0,.5),ylim=c(0,.5))
abline(a=0,b=1,col=2)
```

## Problems: Summary

@@

## FDR via Empirical Bayes

@@

## FDR: A New Deal

@@

## FDR - A New Deal

@@

## Illustration: $g$ a mixture of 0-centered normals

@@

## Illustration: $g$ a mixture of 0-anchored uniforms

@@

## Adaptive Shrinkage

@@

## Recall Problem 1: distribution of alternative Z values multimodal
```{r, echo=FALSE,fig.height=4,fig.cap=""}
nullalthist(hh.zscore,hh.fdrtool$lfdr,main="qvalue")  
```

## Problem Fixed: distribution of alternative Z values unimodal
```{r, echo=FALSE,fig.height=4,fig.cap=""}
nullalthist(hh.zscore,hh.ash$lfdr,main="ash")
```

## Example: FDR estimation

@@

## Recall Problem 2: low count genes add noise, increase q values
```{r,echo=FALSE, fig.height=4,fig.cap=""}
plot(qval.high$qval,qval$qval[highxp],xlab="q values from high-count gene analysis", ylab="q values from all gene analysis", main="q values for high count genes",xlim=c(0,.5),ylim=c(0,.5))
abline(a=0,b=1,col=2)
```

## Problem Fixed: incorporating precision reduces influence of low-count genes
```{r,echo=FALSE, fig.height=4,fig.cap=""}
plot(zdat.ash.high$qvalue,zdat.ash$qvalue[highxp],xlab="q values from high-count gene analysis", ylab="q values from all gene analysis", main="q values for high count genes",xlim=c(0,.5),ylim=c(0,.5))
abline(a=0,b=1,col=2)
```

## A new problem: an embarrassment of riches

@@

## Perhaps we didn't really understand the question?

@@

## The False Sign Rate

@@

## Even with many signals, large $p$ values have high lfsr

@@

## Advantages of False Sign Rate

@@

## Simulated example: $\pi_0$ not identifiable.

@@

## Simulated Example: so fdr not identifiable

@@

## Simulated Example: fsr much more identifiable

@@

## Summary

@@

## Other Applications

@@

## Next steps?

@@

## Thanks

@@

- Including  Chaoxing Dai, Mengyin Lu, Ester Pantaleo, Scott Powers, Sen Tian, Wei Wang, Zhengrong Xing.

@@

## Pandoc Command used

@@
